---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# mwbenchr

<!-- badges: start -->
<!-- badges: end -->

Our mission:

- **To cover all REST API endpoints** provided by the Metabolomics Workbench.
- **To ensure type safety** through strict validation of inputs and outputs.
- **To deliver tidy outputs** as well-structured `data.frame`s ready for analysis.
- **To support efficient workflows** with optional local caching of API responses.
- **To respect API rate limits** with built-in request throttling and retry logic.
- **To enhance user experience** with clear and informative API messages.

Use-case:

```{r}
## ----setup, include=FALSE-----------------------------------------------------
knitr::opts_chunk$set(echo = TRUE)

## ----initialize---------------------------------------------------------------
library(mwbenchr)
client <- mw_rest_client()

## ----compound-----------------------------------------------------------------
# Get compound information by PubChem CID
compound <- get_compound_by_pubchem_cid(client, 5281365)
head(compound)

## ----study--------------------------------------------------------------------
# Get study summary
study <- get_study_summary(client, "ST000001")
head(study)

## ----metstat------------------------------------------------------------------
# Search for diabetes studies in human blood
diabetes_studies <- search_metstat(
  client,
  species = "Human",
  sample_source = "Blood",
  disease = "Diabetes"
)
head(diabetes_studies)
```

